---
title: "Exploratory Data Analysis"
output: github_document
---

## Paul Nguyen & David Herrero-Quevedo

```{r, message = FALSE, cache = TRUE}
#load necessary libraries
library(nbastatR)
library(tidyverse)
library(BBmisc)
```

## Data Description

### Provenance
We obtained our datasets from basketball-reference.com and stats.nba.com.
We used the R package `nbastatR` to obtain the data from those websites.

```{r message = FALSE, cache = TRUE}
players <- bref_players_stats(seasons = 1984:2020, tables = c("advanced", "per_game"), include_all_nba = TRUE)
players_post1984 <- filter(players, yearSeason >= 1984) %>% filter(countGames > 29)
dropping <-c("slugSeason", "idPlayerNBA", "namePlayerBREF", "urlPlayerBREF", "slugPlayerBREF", "urlPlayerThumbnail",
             "urlPlayerHeadshot", "slugPosition","urlPlayerPhoto", "urlPlayerStats", "urlPlayerActionPhoto",
             "yearSeasonFirst","countTeamsPlayerSeason","countGamesStarted","isSeasonCurrent","slugPlayerSeason", "agePlayer",
             "slugTeamBREF", "isHOFPlayer", "slugTeamsBREF")
players_post1984 <- players_post1984[,!(names(players_post1984) %in% dropping)]

for(year in 1984:2020){
  eval(parse(text = paste( "players_",year," <- filter(players_post1984, yearSeason == ",year,")", sep = "" )))
  for(i in 6:48){
    eval(parse(text = paste( "players_",year,"[,",i,"] <- normalize(players_",year,"[,",i,"],method = 'range', range = c(0,1))", sep = "")))
  }
}

players_post1984_normalized <- players_1984
for(year in 1985:2020){
  eval(parse(text = paste( "players_post1984_normalized <- rbind(players_post1984_normalized, players_",year,")", sep = "")))
}
```

### Unit of observation and Variables
Our unit of observation is a player during one season.
Our original dataset consists of 21235 observations  and 75 variables, but we've reduced this number to 
11695 observations after selecting seasons after 1983 and players that have played in at least 30 games in the season.
We've also taken out some unnecessary variables to make a 11695 observation, 55 variable dataset. 
We have included descriptions of the variables at the bottom of the page.

## Data Exploration

### Missingness
The reasoning behind choosing seasons after 1983 is in part that we are not missing any predictor for any observation.

### Univariate analysis of the response
Numerical, graphical (#TODO: need to complete)

### Bi-, trivariate analysis of the response
graphical (#TODO: need to complete)

### Variables used:
#### Shooting:
- pctFTRate
- pct3PRate
- pctFG
- pctFG3
- pctFG2
- pctEFG
- pctFT
- fgmPerGame
- fgaPerGame
- fg3mPerGame
- fg3aPerGame
- fg2mPerGame
- fg2aPerGame
- ftmPerGame
- ftaPerGame
- ptsPerGame

#### Rebounds
- pctORB
- pctTRB
- pctDRB
- orbPerGame
- drbPerGame
- trbPerGame

#### Passing
- pctAST
- pctTOV
- astPerGame
- tovPerGame

#### Defense
- pctSTL
- pctBLK
- stlPerGame
- blkPerGame
- pfPerGame

#### Efficiency stats
- ratioOWS
- ratioDWS
- ratioWS
- ratioWSPer48
- ratioOBPM
- ratioDBPM
- ratioBPM
- ratioVORP

#### Other
- pctUSG
- countTeamsPlayerSeason
- countGamesStarted
- minutesPerGame