---
title: "Exploratory Data Analysis"
output: github_document
---

## Paul Nguyen & David Herrero-Quevedo

```{r, message = FALSE, warning = FALSE, cache = TRUE}
#load necessary libraries
library(nbastatR)
library(tidyverse)
library(BBmisc)
```

## Data Description

### Provenance
We obtained our datasets from basketball-reference.com and stats.nba.com.
We used the R package `nbastatR` to obtain the data from those websites.

```{r message = FALSE, warning = FALSE, cache = TRUE}
players <- bref_players_stats(seasons = 1984:2020, tables = c("advanced", "per_game"), include_all_nba = TRUE)
players_post1984 <- filter(players, yearSeason >= 1984) %>% filter(countGames > 29)
dropping <-c("slugSeason", "idPlayerNBA", "namePlayerBREF", "urlPlayerBREF", "slugPlayerBREF", "urlPlayerThumbnail",
             "urlPlayerHeadshot", "slugPosition","urlPlayerPhoto", "urlPlayerStats", "urlPlayerActionPhoto",
             "yearSeasonFirst","countTeamsPlayerSeason","countGamesStarted","isSeasonCurrent","slugPlayerSeason", "agePlayer",
             "slugTeamBREF", "isHOFPlayer", "slugTeamsBREF")
players_post1984 <- players_post1984[,!(names(players_post1984) %in% dropping)]

for(year in 1984:2020){
  eval(parse(text = paste( "players_",year," <- filter(players_post1984, yearSeason == ",year,")", sep = "" )))
  for(i in 6:48){
    eval(parse(text = paste( "players_",year,"[,",i,"] <- normalize(players_",year,"[,",i,"],method = 'range', range = c(0,1))", sep = "")))
  }
}

players_post1984_normalized <- players_1984
for(year in 1985:2020){
  eval(parse(text = paste( "players_post1984_normalized <- rbind(players_post1984_normalized, players_",year,")", sep = "")))
}
```

### Unit of observation and Variables
Our unit of observation is a player during one season.
Our original dataset consists of 21235 observations  and 75 variables, but we've reduced this number to 
11695 observations after selecting seasons after 1983 and players that have played in at least 30 games in the season.
We've also taken out some unnecessary variables to make a 11695 observation, 55 variable dataset. 
We have included descriptions of the variables at the bottom of the page.

## Data Exploration

### Missingness
The reasoning behind choosing seasons after 1983 is in part that we are not missing any predictor for any observation.

### Univariate analysis of the response
Numerical, graphical (#TODO: need to complete)

### Bi-, trivariate analysis of the response
graphical (#TODO: need to complete)

### Variables used:
#### Shooting:
- pctFTRate: The percentage of total field goal attempts that are Free Throws.
- pct3PRate: The percentage of total field goal attempts that are 3-pt shots.
- pctFG: Field Goal Percentage.
- pctFG3: Field Goal 3-pt Percentage.
- pctFG2: Field Goal 2-pt Percentage.
- pctEFG: Effective Field Goal Percentage. (FGM + (0.5 * 3PM)/FGA). This statistic adjusts for the fact that a 3-pt field goal is worth one more point than a regular 2-pt field goal.
- pctFT: Free Throw Percentage.
- fgmPerGame: Field Goal Made per Game.
- fgaPerGame: Field Goal Attempts per game
- fg3mPerGame: Field Goal Made (3-pt) per game.
- fg3aPerGame: Field Goal Attempts (3-pt) per game.
- fg2mPerGame: Field Goal Made (2-pt) per game.
- fg2aPerGame: Field Goal Attempts (2-pt) per game.
- ftmPerGame: Free Throws Made per game.
- ftaPerGame: Free Throw Attempts per game.
- ptsPerGame: Points per game.

#### Rebounds
- pctORB: Offensive Rebound Percentage (Rebound Rate).
- pctTRB: Total Rebound Percentage (Rebound Rate).
- pctDRB: Deffensive Rebound Percentage (Rebound Rate).
- orbPerGame: Offensive Rebounds per game.
- drbPerGame: Deffensive Rebounds per game.
- trbPerGame: Total Rebounds per game.

#### Passing
- pctAST: Assists Percentage.
- pctTOV: Turnover Percentage.
- astPerGame: Assists per game.
- tovPerGame: Turnovers per game.

#### Defense
- pctSTL: Steals percentage.
- pctBLK: Blocks percentage.
- stlPerGame: Steals per game.
- blkPerGame: Blocks per game.
- pfPerGame: Personal Fouls per game.

#### Efficiency stats
- ratioOWS: Offensive Win Shares.
- ratioDWS: Deffensive Win Shares.
- ratioWS: Win Shares.
- ratioWSPer48: Win Shares per 48 minutes.
- ratioOBPM: Offensive Box Plus/Minus.
- ratioDBPM: Deffensive Box Plus/Minus.
- ratioBPM: Box Plus/Minus.
- ratioVORP: Value Over Replacement Player.
- minutesPerGame: Minutes Per Game
- pctUSG: Usage Percentage. (100 * ((FGA + 0.44 * FTA + TOV) * (Tm MP / 5)) / (MP * (Tm FGA + 0.44 * Tm FTA + Tm TOV))
