---
title: "Exploratory Data Analysis"
output: github_document
---

## Paul Nguyen & David Herrero-Quevedo

```{r, message = FALSE, cache = TRUE}
#load necessary libraries
library(ggplot2)
library(dplyr)
library(nbastatR)
library(tidyr)
library(tidyverse)
library(BBmisc)
```


## Data Description

### Provenance
We obtained our datasets from basketball-reference.com and stats.nba.com.
We used the R package `nbastatR` to obtain the data from those websites.

### Unit of observation and Variables
Our unit of observation is a player during one season.
Our original dataset consists of 21235 observations  and 75 variables, but 
we've reduced this number to 10223 observations after selecting seasons
after 1983 and players that have played at least 600 minutes in the season.
We've also taken out some unnecessary variables to make a 14528 observation,
64 variable dataset.

```{r, message = FALSE, cache = TRUE}
#grabbing and filtering data
players <- bref_players_stats(seasons = 1950:2020, tables = c("advanced", "per_game"), include_all_nba = TRUE)
players_post1984 <- filter(players, yearSeason >= 1984) 
ggplot(players_post1984, mapping = aes(x = minutes)) + 
  geom_histogram() +
  geom_vline(xintercept = 600)
players_post1984 <- filter(players_post1984, minutes > 600)

#getting rid of some not useful variables
drops <- c("slugSeason", "idPlayerNBA", "namePlayerBREF", 
           "slugPlayerBREF", "urlPlayerThumbnail", "urlPlayerHeadshot",
           "urlPlayerPhoto", "urlPlayerStats", "urlPlayerActionPhoto",
           "isSeasonCurrent","slugPlayerSeason")

players_post1984 <-
  players_post1984[,!(names(players_post1984) %in% drops)]

```



## Data Exploration

```{r}
#some summary statistics
summarystatistics1984 <- players_post1984 %>%
  group_by(yearSeason, groupPosition) %>%
  summarize(meanpctFTRate = mean(pctFTRate),
            meanpct3PRate = mean(pct3PRate),
            meanpctFG = mean(pctFG),
            meanpctFG3 = mean(pctFG3),
            meanpctFG2 = mean(pctFG2),
            meanpctEFG = mean(pctEFG),
            meanpctFT = mean(pctFT),
            meanfgmPerGame = mean(fgmPerGame),
            meanfgaPerGame = mean(fgaPerGame),
            meanfg3mPerGame = mean(fg3mPerGame),
            meanfg3aPerGame = mean(fg3aPerGame),
            meanfg2mPerGame = mean(fg2mPerGame),
            meanfg2aPerGame = mean(fg2aPerGame),
            meanftmPerGame = mean(ftmPerGame),
            meanftaPerGame = mean(ftaPerGame),
            meanptsPerGame = mean(ptsPerGame),
            meanpctORB = mean(pctORB),
            meanpctTRB = mean(pctTRB),
            meanpctDRB = mean(pctDRB),
            meanorbPerGame = mean(orbPerGame),
            meandrbPerGame = mean (drbPerGame),
            meantrbPerGame = mean(trbPerGame),
            meanpctAST = mean(pctAST),
            meanpctTOV = mean(pctTOV),
            meanastPerGame = mean(astPerGame),
            meantovPerGame = mean(tovPerGame),
            meanpctSTL = mean(pctSTL),
            meanpctBLK = mean(pctBLK),
            meanstlPerGame = mean(stlPerGame),
            meanblkPerGame = mean(blkPerGame),
            meanpfPerGame = mean(pfPerGame),
            meanratioOWS = mean(ratioOWS),
            meanratioDWS = mean(ratioDWS),
            meanratioWS = mean(ratioWS),
            meanratioWSPer48 = mean(ratioWSPer48),
            meanratioOBPM = mean(ratioOBPM),
            meanratioDBPM = mean(ratioDBPM),
            meanratioBPM = mean(ratioBPM),
            meanratioVORP = mean(ratioVORP),
            sdpctFTRate = sd(pctFTRate),
            sdpct3PRate = sd(pct3PRate),
            sdpctFG = sd(pctFG),
            sdpctFG3 = sd(pctFG3),
            sdpctFG2 = sd(pctFG2),
            sdpctEFG = sd(pctEFG),
            sdpctFT = sd(pctFT),
            sdfgmPerGame = sd(fgmPerGame),
            sdfgaPerGame = sd(fgaPerGame),
            sdfg3mPerGame = sd(fg3mPerGame),
            sdfg3aPerGame = sd(fg3aPerGame),
            sdfg2mPerGame = sd(fg2mPerGame),
            sdfg2aPerGame = sd(fg2aPerGame),
            sdftmPerGame = sd(ftmPerGame),
            sdftaPerGame = sd(ftaPerGame),
            sdptsPerGame = sd(ptsPerGame),
            sdpctORB = sd(pctORB),
            sdpctTRB = sd(pctTRB),
            sdpctDRB = sd(pctDRB),
            sdorbPerGame = sd(orbPerGame),
            sddrbPerGame = sd (drbPerGame),
            sdtrbPerGame = sd(trbPerGame),
            sdpctSTL = sd(pctSTL),
            sdpctBLK = sd(pctBLK),
            sdstlPerGame = sd(stlPerGame),
            sdblkPerGame = sd(blkPerGame),
            sdpfPerGame = sd(pfPerGame),
            sdratioOWS = sd(ratioOWS),
            sdratioDWS = sd(ratioDWS),
            sdratioWS = sd(ratioWS),
            sdratioWSPer48 = sd(ratioWSPer48),
            sdratioOBPM = sd(ratioOBPM),
            sdratioDBPM = sd(ratioDBPM),
            sdratioBPM = sd(ratioBPM),
            sdratioVORP = sd(ratioVORP)
            )
head(summarystatistics1984)

ggplot(data = summarystatistics1984, mapping = aes(x = yearSeason, y = meanpctFG, color = groupPosition)) + 
  geom_line()

#taking a look at some stats over time, grouped by year and position
#fg pct
ggplot(data = summarystatistics1984, 
       mapping = aes(x = yearSeason, y = meanpctFG, color = groupPosition)) +  geom_line()
#3FGPct, see the introduction of 3 point line
ggplot(data = summarystatistics1984, 
       mapping = aes(x = yearSeason, y = meanpctFG3,
                     color = groupPosition)) +
  geom_line()
#meanFGM
ggplot(data = summarystatistics1984, 
       mapping = aes(x = yearSeason, y = meanfgmPerGame,
                     color = groupPosition)) +
  geom_line()
#meanFGA
ggplot(data = summarystatistics1984, 
       mapping = aes(x = yearSeason, y = meanfgaPerGame,
                     color = groupPosition)) +
  geom_line()

#mean3FGM 
ggplot(data = summarystatistics1984, 
       mapping = aes(x = yearSeason, y = meanfg3mPerGame,
                     color = groupPosition)) +
  geom_line()
#mean3FGA
ggplot(data = summarystatistics1984, 
       mapping = aes(x = yearSeason, y = meanfg3aPerGame, color = groupPosition)) +
  geom_line()
#meanFTA
ggplot(data = summarystatistics1984, 
       mapping = aes(x = yearSeason, y = meanftaPerGame, 
                     color = groupPosition)) +
  geom_line()
#meanFTM
ggplot(data = summarystatistics1984, 
       mapping = aes(x = yearSeason, y = meanftmPerGame,
                     color = groupPosition)) +
  geom_line()
#meanORB
ggplot(data = summarystatistics1984, 
       mapping = aes(x = yearSeason, y = meanorbPerGame,
                     color = groupPosition)) +
  geom_line()
#meanDRB
ggplot(data = summarystatistics1984, 
       mapping = aes(x = yearSeason, y = meandrbPerGame,
                     color = groupPosition)) +
  geom_line()
#meanAST
ggplot(data = summarystatistics1984, 
       mapping = aes(x = yearSeason, y = meanastPerGame, 
                     color = groupPosition)) +
  geom_line()
#meanSTL
ggplot(data = summarystatistics1984, 
       mapping = aes(x = yearSeason, y = meanstlPerGame,
                     color = groupPosition)) +
  geom_line()
#meanBLK
ggplot(data = summarystatistics1984, 
       mapping = aes(x = yearSeason, y = meanblkPerGame,
                     color = groupPosition)) +
  geom_line()
#meanTOV
ggplot(data = summarystatistics1984, 
       mapping = aes(x = yearSeason, y = meantovPerGame, 
                     color = groupPosition)) +
  geom_line()
#meanPTS
ggplot(data = summarystatistics1984, 
       mapping = aes(x = yearSeason, y = meanptsPerGame,
                     color = groupPosition)) +
  geom_line()


```


### Missingness
(#TODO: need to complete)

### Univariate analysis of the response
Numerical, graphical (#TODO: need to complete)

### Bi-, trivariate analysis of the response
graphical (#TODO: need to complete)

### Variables used:
#### Shooting:
- pctFTRate
- pct3PRate
- pctFG
- pctFG3
- pctFG2
- pctEFG
- pctFT
- fgmPerGame
- fgaPerGame
- fg3mPerGame
- fg3aPerGame
- fg2mPerGame
- fg2aPerGame
- ftmPerGame
- ftaPerGame
- ptsPerGame

#### Rebounds
- pctORB
- pctTRB
- pctDRB
- orbPerGame
- drbPerGame
- trbPerGame

#### Passing
- pctAST
- pctTOV
- astPerGame
- tovPerGame

#### Defense
- pctSTL
- pctBLK
- stlPerGame
- blkPerGame
- pfPerGame

#### Efficiency stats
- ratioOWS
- ratioDWS
- ratioWS
- ratioWSPer48
- ratioOBPM
- ratioDBPM
- ratioBPM
- ratioVORP

#### Other
- pctUSG
- countTeamsPlayerSeason
- countGamesStarted
- minutesPerGame