---
title: "Exploratory Data Analysis"
output: github_document
---

## Paul Nguyen & David Herrero-Quevedo

```{r, message = FALSE, warning = FALSE, cache = TRUE}
#load necessary libraries
library(nbastatR)
library(tidyverse)
library(BBmisc)
library(ggplot2)
library(dplyr)
```

## Data Description

### Provenance
We obtained our datasets from basketball-reference.com and stats.nba.com.
We used the R package `nbastatR` to obtain the data from those websites.

```{r message = FALSE, warning = FALSE, cache = TRUE}
# Gets player stats from 1984 to 2020.
players_post1984 <- bref_players_stats(seasons = 1984:2020, tables = c("advanced", "per_game"), include_all_nba = TRUE)

# Filters by amount of time they played.
ggplot(players_post1984, mapping = aes(x = minutes)) + 
   geom_histogram() +
   geom_vline(xintercept = 600)
 players_post1984 <- filter(players_post1984, minutes > 600)

 # Variables to drop
dropping <-c("slugSeason", "idPlayerNBA", "namePlayerBREF", "urlPlayerBREF",
             "slugPlayerBREF", "urlPlayerThumbnail", "urlPlayerHeadshot",
             "slugPosition","urlPlayerPhoto", "urlPlayerStats",
             "urlPlayerActionPhoto","yearSeasonFirst",
             "countTeamsPlayerSeason","countGamesStarted",
             "isSeasonCurrent","slugPlayerSeason", "agePlayer",
             "slugTeamBREF", "isHOFPlayer", "slugTeamsBREF", 
             "countTeamsPlayerSeasonPerGame", "groupAllNBA",
             "numberAllNBATeam", "isAllNBA2", "isAllNBA1",
             "isAllNBA3" )
# Dropping variables
players_post1984 <- players_post1984[,!(names(players_post1984) %in% dropping)]

# Groups players by season, and normalizes by season.
for(year in 1984:2020){
  eval(parse(text = paste( "players_",year," <- filter(players_post1984, yearSeason == ",year,")", sep = "" )))
  for(i in 6:48){
    eval(parse(text = paste( "players_",year,"[,",i,"] <- normalize(players_",year,"[,",i,"],method = 'range', range = c(0,1))", sep = "")))
  }
}

# Ungroups players by season.
players_post1984_normalized <- players_1984
for(year in 1985:2020){
  eval(parse(text = paste( "players_post1984_normalized <- rbind(players_post1984_normalized, players_",year,")", sep = "")))
}
```

### Unit of observation and Variables
Our unit of observation is a player during one season.
Our original dataset consists of 21235 observations  and 75 variables, but we've reduced this number to 
10223 observations after selecting seasons after 1983 and players that have played in at least 600 minutes in the season.
We've also taken out some unnecessary variables to make a 11695 observation, 49 variable dataset. 
We have included descriptions of the variables at the bottom of the page.

## Data Exploration

```{r}
#some summary statistics
summarystatistics1984 <- players_post1984 %>%
  group_by(yearSeason, groupPosition) %>%
  summarize(meanpctFTRate = mean(pctFTRate),
            meanpct3PRate = mean(pct3PRate),
            meanpctFG = mean(pctFG),
            meanpctFG3 = mean(pctFG3),
            meanpctFG2 = mean(pctFG2),
            meanpctEFG = mean(pctEFG),
            meanpctFT = mean(pctFT),
            meanfgmPerGame = mean(fgmPerGame),
            meanfgaPerGame = mean(fgaPerGame),
            meanfg3mPerGame = mean(fg3mPerGame),
            meanfg3aPerGame = mean(fg3aPerGame),
            meanfg2mPerGame = mean(fg2mPerGame),
            meanfg2aPerGame = mean(fg2aPerGame),
            meanftmPerGame = mean(ftmPerGame),
            meanftaPerGame = mean(ftaPerGame),
            meanptsPerGame = mean(ptsPerGame),
            meanpctORB = mean(pctORB),
            meanpctTRB = mean(pctTRB),
            meanpctDRB = mean(pctDRB),
            meanorbPerGame = mean(orbPerGame),
            meandrbPerGame = mean (drbPerGame),
            meantrbPerGame = mean(trbPerGame),
            meanpctAST = mean(pctAST),
            meanpctTOV = mean(pctTOV),
            meanastPerGame = mean(astPerGame),
            meantovPerGame = mean(tovPerGame),
            meanpctSTL = mean(pctSTL),
            meanpctBLK = mean(pctBLK),
            meanstlPerGame = mean(stlPerGame),
            meanblkPerGame = mean(blkPerGame),
            meanpfPerGame = mean(pfPerGame),
            meanratioOWS = mean(ratioOWS),
            meanratioDWS = mean(ratioDWS),
            meanratioWS = mean(ratioWS),
            meanratioWSPer48 = mean(ratioWSPer48),
            meanratioOBPM = mean(ratioOBPM),
            meanratioDBPM = mean(ratioDBPM),
            meanratioBPM = mean(ratioBPM),
            meanratioVORP = mean(ratioVORP),
            sdpctFTRate = sd(pctFTRate),
            sdpct3PRate = sd(pct3PRate),
            sdpctFG = sd(pctFG),
            sdpctFG3 = sd(pctFG3),
            sdpctFG2 = sd(pctFG2),
            sdpctEFG = sd(pctEFG),
            sdpctFT = sd(pctFT),
            sdfgmPerGame = sd(fgmPerGame),
            sdfgaPerGame = sd(fgaPerGame),
            sdfg3mPerGame = sd(fg3mPerGame),
            sdfg3aPerGame = sd(fg3aPerGame),
            sdfg2mPerGame = sd(fg2mPerGame),
            sdfg2aPerGame = sd(fg2aPerGame),
            sdftmPerGame = sd(ftmPerGame),
            sdftaPerGame = sd(ftaPerGame),
            sdptsPerGame = sd(ptsPerGame),
            sdpctORB = sd(pctORB),
            sdpctTRB = sd(pctTRB),
            sdpctDRB = sd(pctDRB),
            sdorbPerGame = sd(orbPerGame),
            sddrbPerGame = sd (drbPerGame),
            sdtrbPerGame = sd(trbPerGame),
            sdpctSTL = sd(pctSTL),
            sdpctBLK = sd(pctBLK),
            sdstlPerGame = sd(stlPerGame),
            sdblkPerGame = sd(blkPerGame),
            sdpfPerGame = sd(pfPerGame),
            sdratioOWS = sd(ratioOWS),
            sdratioDWS = sd(ratioDWS),
            sdratioWS = sd(ratioWS),
            sdratioWSPer48 = sd(ratioWSPer48),
            sdratioOBPM = sd(ratioOBPM),
            sdratioDBPM = sd(ratioDBPM),
            sdratioBPM = sd(ratioBPM),
            sdratioVORP = sd(ratioVORP)
            )
head(summarystatistics1984)
ggplot(data = summarystatistics1984, mapping = aes(x = yearSeason, y = meanpctFG, color = groupPosition)) + 
  geom_line()
#taking a look at some stats over time, grouped by year and position
#fg pct
ggplot(data = summarystatistics1984, 
       mapping = aes(x = yearSeason, y = meanpctFG, color = groupPosition)) +  geom_line()
#3FGPct, see the introduction of 3 point line
ggplot(data = summarystatistics1984, 
       mapping = aes(x = yearSeason, y = meanpctFG3,
                     color = groupPosition)) +
  geom_line()
#meanFGM
ggplot(data = summarystatistics1984, 
       mapping = aes(x = yearSeason, y = meanfgmPerGame,
                     color = groupPosition)) +
  geom_line()
#meanFGA
ggplot(data = summarystatistics1984, 
       mapping = aes(x = yearSeason, y = meanfgaPerGame,
                     color = groupPosition)) +
  geom_line()
#mean3FGM 
ggplot(data = summarystatistics1984, 
       mapping = aes(x = yearSeason, y = meanfg3mPerGame,
                     color = groupPosition)) +
  geom_line()
#mean3FGA
ggplot(data = summarystatistics1984, 
       mapping = aes(x = yearSeason, y = meanfg3aPerGame, color = groupPosition)) +
  geom_line()
#meanFTA
ggplot(data = summarystatistics1984, 
       mapping = aes(x = yearSeason, y = meanftaPerGame, 
                     color = groupPosition)) +
  geom_line()
#meanFTM
ggplot(data = summarystatistics1984, 
       mapping = aes(x = yearSeason, y = meanftmPerGame,
                     color = groupPosition)) +
  geom_line()
#meanORB
ggplot(data = summarystatistics1984, 
       mapping = aes(x = yearSeason, y = meanorbPerGame,
                     color = groupPosition)) +
  geom_line()
#meanDRB
ggplot(data = summarystatistics1984, 
       mapping = aes(x = yearSeason, y = meandrbPerGame,
                     color = groupPosition)) +
  geom_line()
#meanAST
ggplot(data = summarystatistics1984, 
       mapping = aes(x = yearSeason, y = meanastPerGame, 
                     color = groupPosition)) +
  geom_line()
#meanSTL
ggplot(data = summarystatistics1984, 
       mapping = aes(x = yearSeason, y = meanstlPerGame,
                     color = groupPosition)) +
  geom_line()
#meanBLK
ggplot(data = summarystatistics1984, 
       mapping = aes(x = yearSeason, y = meanblkPerGame,
                     color = groupPosition)) +
  geom_line()
#meanTOV
ggplot(data = summarystatistics1984, 
       mapping = aes(x = yearSeason, y = meantovPerGame, 
                     color = groupPosition)) +
  geom_line()
#meanPTS
ggplot(data = summarystatistics1984, 
       mapping = aes(x = yearSeason, y = meanptsPerGame,
                     color = groupPosition)) +
  geom_line()
```

### Missingness
The reasoning behind choosing seasons after 1983 is in part so that we are
not missing any predictors for any observation. For example, the three 
point line was introduced in the 79/80 season.


### Variables used:
#### Shooting:
- pctFTRate: The percentage of total field goal attempts that are Free Throws.
- pct3PRate: The percentage of total field goal attempts that are 3-pt shots.
- pctFG: Field Goal Percentage.
- pctFG3: Field Goal 3-pt Percentage.
- pctFG2: Field Goal 2-pt Percentage.
- pctEFG: Effective Field Goal Percentage. (FGM + (0.5 * 3PM)/FGA). This statistic adjusts for the fact that a 3-pt field goal is worth one more point than a regular 2-pt field goal.
- pctFT: Free Throw Percentage.
- fgmPerGame: Field Goal Made per Game.
- fgaPerGame: Field Goal Attempts per game
- fg3mPerGame: Field Goal Made (3-pt) per game.
- fg3aPerGame: Field Goal Attempts (3-pt) per game.
- fg2mPerGame: Field Goal Made (2-pt) per game.
- fg2aPerGame: Field Goal Attempts (2-pt) per game.
- ftmPerGame: Free Throws Made per game.
- ftaPerGame: Free Throw Attempts per game.
- ptsPerGame: Points per game.

#### Rebounds
- pctORB: Offensive Rebound Percentage (Rebound Rate).
- pctTRB: Total Rebound Percentage (Rebound Rate).
- pctDRB: Deffensive Rebound Percentage (Rebound Rate).
- orbPerGame: Offensive Rebounds per game.
- drbPerGame: Deffensive Rebounds per game.
- trbPerGame: Total Rebounds per game.

#### Passing
- pctAST: Assists Percentage.
- pctTOV: Turnover Percentage.
- astPerGame: Assists per game.
- tovPerGame: Turnovers per game.

#### Defense
- pctSTL: Steals percentage.
- pctBLK: Blocks percentage.
- stlPerGame: Steals per game.
- blkPerGame: Blocks per game.
- pfPerGame: Personal Fouls per game.

#### Efficiency stats
- ratioPER: Player Efficiency Rating
- ratioOWS: Offensive Win Shares.
- ratioDWS: Deffensive Win Shares.
- ratioWS: Win Shares.
- ratioWSPer48: Win Shares per 48 minutes.
- ratioOBPM: Offensive Box Plus/Minus.
- ratioDBPM: Deffensive Box Plus/Minus.
- ratioBPM: Box Plus/Minus.
- ratioVORP: Value Over Replacement Player.
- minutesPerGame: Minutes Per Game
- pctUSG: Usage Percentage. (100 * ((FGA + 0.44 * FTA + TOV) * (Tm MP / 5)) / (MP * (Tm FGA + 0.44 * Tm FTA + Tm TOV))


## Looking at PCA (Principal Component Analysis)

```{r}
pca <- prcomp(players_post1984_normalized[,-c(1:5, 49)])
d <- as.data.frame(pca$x)
pc1 <- pca$rotation[, 1]
pc1
pc2 <- pca$rotation[, 2]
pc2
d$namePlayer <- players_post1984_normalized$namePlayer
d$isAllNBA <- players_post1984_normalized$isAllNBA
d$yearSeason <- players_post1984_normalized$yearSeason

pcaplot <- ggplot(d, aes(x = PC1, y = PC2)) +
  geom_point(size = .5, alpha = .7) 
pcaplot

d1 <- data.frame(PC = 1:43,
                PVE = pca$sdev^2 /
                  sum(pca$sdev^2))

ggplot(d1, aes(x = PC, y = PVE)) +
  geom_line() + 
  geom_point()






#lets now create a plot with some players that we will recognize
currentplayersplot <- ggplot(d, aes(x = PC1, y = PC2)) +
  geom_point(size = .5, alpha = .1) +
  geom_text(data = subset(d, isAllNBA == TRUE & 
                            yearSeason %in% 2017:2019 ),
            aes(label = namePlayer))
currentplayersplot

famousplayersplot <-  ggplot(d, aes(x = PC1, y = PC2)) +
  geom_point(size = .5, alpha = .1) +
  geom_text(data = subset(d, namePlayer %in% c("Lebron James", 
                                               "Michael Jordan",
                                               "Stephen Curry",
                                               "Kobe Bryant",
                                               "Tim Duncan",
                                               "Magic Johnson",
                                               "Kevin Durant",
                                               "Corey Maggette")),
            aes(label = namePlayer))
famousplayersplot
#corey maguette is included here because i remember his name being 
#funny when i watched basketball
```
PC1 seems to be a case of an "Usage" (-) vs "Non-Usage" (+) battle.
In the negative side, we see characteristics such as ptsPerGame,
fgmPerGame, fg2mPerGame, fg2aPerGame. The other types of variables 
do not tend to take on positive values, but the ones that are include:
pctBLK, pctSTL, pctTOV.

Looking at PC2, I would describe this as "Shooters" (-) vs "Baseline" (+).
We see negative values for shooting characteristics, such as pct3PRate, 
fg3mPerGame, fg3aPerGame, and positive values for characteristics 
typical for tall, big players:pctDRB, pctTRB, blkPerGame
