---
title: "NBA Modern Positions"
output: github_document
---

```{r,warning = FALSE, message = FALSE}
#load necessary libraries
library(nbastatR)
library(tidyverse)
library(BBmisc)
library(corrplot)
library(gganimate)
```

## Abstract

## Introduction

### The data

#### Provenance

We obtained our datasets from basketball-reference.com and stats.nba.com.
We used the R package `nbastatR` to obtain the data from those websites.

#### Unit of observation and Variables

Our unit of observation is a player during one season.
Our original dataset consists of 21235 observations  and 75 variables, but we've reduced this number to 
10223 observations after selecting seasons after 1983 and players that have played in at least 600 minutes in the season.
We've also taken out some unnecessary variables to make a 11695 observation, 49 variable dataset. 
We have included descriptions of the variables at the bottom of the page.

```{r message = FALSE, warning = FALSE, cache = TRUE}
# Gets player stats from 1984 to 2020.
players_post1984 <- bref_players_stats(seasons = 1984:2020, tables = c("advanced", "per_game"), include_all_nba = TRUE, return_message = FALSE)

# Filters by amount of time they played.
ggplot(players_post1984, mapping = aes(x = minutes)) + 
   geom_histogram() +
   geom_vline(xintercept = 600)
 players_post1984 <- filter(players_post1984, minutes > 600)

 # Variables to drop
dropping <-c("slugSeason", "idPlayerNBA", "namePlayerBREF", "urlPlayerBREF",
             "slugPlayerBREF", "urlPlayerThumbnail", "urlPlayerHeadshot",
             "slugPosition","urlPlayerPhoto", "urlPlayerStats",
             "urlPlayerActionPhoto","yearSeasonFirst",
             "countTeamsPlayerSeason","countGamesStarted",
             "isSeasonCurrent","slugPlayerSeason", "agePlayer",
             "slugTeamBREF", "isHOFPlayer", "slugTeamsBREF", 
             "countTeamsPlayerSeasonPerGame", "groupAllNBA",
             "numberAllNBATeam", "isAllNBA2", "isAllNBA1",
             "isAllNBA3" )
# Dropping variables
players_post1984 <- players_post1984[,!(names(players_post1984) %in% dropping)]

# Groups players by season, and normalizes by season.
for(year in 1984:2020){
  eval(parse(text = paste( "players_",year," <- filter(players_post1984, yearSeason == ",year,")", sep = "" )))
  for(i in 6:48){
    eval(parse(text = paste( "players_",year,"[,",i,"] <- normalize(players_",year,"[,",i,"],method = 'range', range = c(0,1))", sep = "")))
  }
}

# Ungroups players by season.
players_post1984_normalized <- players_1984
for(year in 1985:2020){
  eval(parse(text = paste( "players_post1984_normalized <- rbind(players_post1984_normalized, players_",year,")", sep = "")))
}
```

### Variables used:
#### Shooting:
- pctFTRate: The percentage of total field goal attempts that are Free Throws.
- pct3PRate: The percentage of total field goal attempts that are 3-pt shots.
- pctFG: Field Goal Percentage.
- pctFG3: Field Goal 3-pt Percentage.
- pctFG2: Field Goal 2-pt Percentage.
- pctEFG: Effective Field Goal Percentage. (FGM + (0.5 * 3PM)/FGA). This statistic adjusts for the fact that a 3-pt field goal is worth one more point than a regular 2-pt field goal.
- pctFT: Free Throw Percentage.
- fgmPerGame: Field Goal Made per Game.
- fgaPerGame: Field Goal Attempts per game
- fg3mPerGame: Field Goal Made (3-pt) per game.
- fg3aPerGame: Field Goal Attempts (3-pt) per game.
- fg2mPerGame: Field Goal Made (2-pt) per game.
- fg2aPerGame: Field Goal Attempts (2-pt) per game.
- ftmPerGame: Free Throws Made per game.
- ftaPerGame: Free Throw Attempts per game.
- ptsPerGame: Points per game.

#### Rebounds
- pctORB: Offensive Rebound Percentage (Rebound Rate).
- pctTRB: Total Rebound Percentage (Rebound Rate).
- pctDRB: Deffensive Rebound Percentage (Rebound Rate).
- orbPerGame: Offensive Rebounds per game.
- drbPerGame: Deffensive Rebounds per game.
- trbPerGame: Total Rebounds per game.

#### Passing
- pctAST: Assists Percentage.
- pctTOV: Turnover Percentage.
- astPerGame: Assists per game.
- tovPerGame: Turnovers per game.

#### Defense
- pctSTL: Steals percentage.
- pctBLK: Blocks percentage.
- stlPerGame: Steals per game.
- blkPerGame: Blocks per game.
- pfPerGame: Personal Fouls per game.

#### Efficiency stats
- ratioPER: Player Efficiency Rating
- ratioOWS: Offensive Win Shares.
- ratioDWS: Deffensive Win Shares.
- ratioWS: Win Shares.
- ratioWSPer48: Win Shares per 48 minutes.
- ratioOBPM: Offensive Box Plus/Minus.
- ratioDBPM: Deffensive Box Plus/Minus.
- ratioBPM: Box Plus/Minus.
- ratioVORP: Value Over Replacement Player.
- minutesPerGame: Minutes Per Game
- pctUSG: Usage Percentage. (100 * ((FGA + 0.44 * FTA + TOV) * (Tm MP / 5)) / (MP * (Tm FGA + 0.44 * Tm FTA + Tm TOV))

## Modeling

```{r}
pca <- prcomp(players_post1984_normalized[,-c(1:5, 49)])
d <- as.data.frame(pca$x)
pc1 <- pca$rotation[, 1]
pc1
pc2 <- pca$rotation[, 2]
pc2
pca3 <- pca$rotation[, 3]
d$namePlayer <- players_post1984_normalized$namePlayer
d$isAllNBA <- players_post1984_normalized$isAllNBA
d$yearSeason <- players_post1984_normalized$yearSeason

pcaplot <- ggplot(d, aes(x = PC1, y = PC2)) +
  geom_point(size = .5, alpha = .7) +
  ylab("Shooters vs Baseline") +
  xlab("Usage vs Non-Usage")
pcaplot

d1 <- data.frame(PC = 1:43,
                PVE = pca$sdev^2 /
                  sum(pca$sdev^2))

ggplot(d1, aes(x = PC, y = PVE)) +
  geom_line() + 
  geom_point()

#some PCA exploration
pcarotations <- data.frame(pca$rotation)
pcarotations$variables <- rownames(pcarotations)
PC1graph <- 
  ggplot(data = pcarotations, mapping = aes(x = variables, y = PC1, 
                                          fill = variables)) +
  geom_col() +
  coord_flip() +
  theme(legend.position = "none") +
  labs(title = "Usage (-) vs Non-Usage (+)")
PC1graph
#note important characteristics: tov, pts, minutes, fgm, fga
PC2graph <- 
  ggplot(data = pcarotations, mapping = aes(x = variables, y = PC2, 
                                          fill = variables)) +
  geom_col() +
  coord_flip() +
  theme(legend.position = "none")+
  labs(title = "Shooters vs Baseline")
PC2graph
#note important characteristics: (+) rebounding stats, (-) 3pt shooting, asts
PC3graph <- 
  ggplot(data = pcarotations, mapping = aes(x = variables, y = PC3, 
                                          fill = variables)) +
  geom_col() +
  coord_flip() +
  theme(legend.position = "none")
PC3graph
#idk this one seems pretty hard to interpret lol

#lets now create a plot with some players that we will recognize
currentplayersplot <- ggplot(d, aes(x = PC1, y = PC2)) +
  geom_point(size = .5, alpha = .1) +
  geom_text(data = subset(d, isAllNBA == TRUE & 
                            yearSeason %in% 2017:2019 ),
            aes(label = namePlayer)) +
  ylab("Shooters vs Baseline") +
  xlab("Usage vs Non-Usage")
currentplayersplot

famousplayersplot <-  ggplot(d, aes(x = PC1, y = PC2)) +
  geom_point(size = .5, alpha = .1) +
  geom_text(data = subset(d, namePlayer %in% c("Lebron James", 
                                               "Michael Jordan",
                                               "Stephen Curry",
                                               "Kobe Bryant",
                                               "Tim Duncan",
                                               "Magic Johnson",
                                               "Kevin Durant")),
            aes(label = namePlayer)) +
  geom_point(color = "purple",
             data = subset(d, namePlayer %in% c("Lebron James", 
                                               "Michael Jordan",
                                               "Stephen Curry",
                                               "Kobe Bryant",
                                               "Tim Duncan",
                                               "Magic Johnson",
                                               "Kevin Durant"),
             size = .5, alpha = .8)) +
  ylab("Shooters vs Baseline") +
  xlab("Usage vs Non-Usage")
famousplayersplot
#animated plot, allstars throughout the seasons
allnbaplayerspca <- d %>%
  filter(isAllNBA == TRUE)
#staticplot
staticplot <- ggplot(data = allnbaplayerspca, 
                     mapping = aes(x = PC1, y = PC2)) +
  geom_point(alpha = .8, color = "purple") + 
  geom_text(data = allnbaplayerspca, mapping = aes(label = namePlayer)) +
  geom_point(data = d, mapping = aes(x = PC1, y = PC2), alpha = .1) +
  ylab("Shooters vs Baseline") +
  xlab("Usage vs Non-Usage")
staticplot


#byseason
byseasonplot <- staticplot + transition_time(yearSeason) +
  labs(title = "Season: {frame_time}")
animate(byseasonplot, renderer = gifski_renderer(), nframes = 36, fps = 2)

```

```{r}
#to do: k-means clustering, hierchical clustering
n_pcas <- as_tibble(pca$x[,1:3])
n_clusters_kmeans<-kmeans(n_pcas, centers = 8)
n_clusters_hier<-hclust(dist(n_pcas))
n_clusters_hier_plot<-plot(n_clusters_hier)

n_clusters_hier_cut <- cutree(n_clusters_hier, 10)

n_pcas_kmeans <- n_pcas %>% mutate(cluster = n_clusters_kmeans$cluster)
n_pcas_hier <- n_pcas %>% mutate(cluster = n_clusters_hier_cut)

ggplot(n_pcas_kmeans, aes(x = PC1, y = PC2, color = cluster)) + geom_point()
ggplot(n_pcas_hier, aes(x = PC1, y = PC2, color = cluster)) + geom_point()

```

## Discussion

### Looking at PCA (Principal Component Analysis)

PC1 seems to be a case of an "Usage" (-) vs "Non-Usage" (+) battle.
In the negative side, we see characteristics such as ptsPerGame,
fgmPerGame, fg2mPerGame, fg2aPerGame. The other types of variables 
do not tend to take on positive values, but the ones that are include:
pctBLK, pctSTL, pctTOV.  

Looking at PC2, I would describe this as "Shooters" (-) vs "Baseline" (+).
We see negative values for shooting characteristics, such as pct3PRate, 
fg3mPerGame, fg3aPerGame, and positive values for characteristics 
typical for tall, big players:pctDRB, pctTRB, blkPerGame.

## References