---
title: "NBA Modern Positions"
output: github_document
always_allow_html: true
---

```{r,warning = FALSE, message = FALSE}
#load necessary libraries
library(nbastatR)
library(tidyverse)
library(BBmisc)
library(corrplot)
library(gganimate)
library(plotly)
library(ggrepel)
```

## Abstract

In this project, my group will describe the changes in playstyle types
throughout the NBA's history. We will perform PCA analysis to determine
the essential traits for different players, and then use important principal
components and unsupervised learning techinques to cluster players into
different "types" and see the movement of types over time. 

## Introduction

The NBA, initially created in 1946, has seen much change throughout its 
history, for example, the introduction of the 3 point line in the 79/80
season and the elimination of hand checking in 1990. These changes have
altered the playstyles of many teams; some players, especially 
former players, have critisized these changes, such as the reliance of many
teams on jump shooting. If you look at the Houston Rocket's shotchart,
you may notice huge regions of white space away from the basket and the 
3 point line in order to maximize the efficiency of their shots. We expect
to see a steady increase in 3 point shooting from when the 3 point line 
was introduced in the NBA. We also expect to see a decrease in big, slow
players that may have trouble defending these shots, and a decrease in 
players that specialize in shooting the midrange.

### The data

We obtained our datasets from basketball-reference.com and stats.nba.com.
We used the R package `nbastatR` to obtain the data from those websites.

#### Unit of observation and Variables

Our unit of observation is a player during one season.
Our original dataset consists of 21235 observations  and 75 variables, but we've reduced this number to  10223 observations after selecting seasonsafter 1983 and players that have played in at least 600 minutes in the season. We've also taken out some unnecessary variables to make a 11695 observation, 49 variable dataset. 

### Variables used:
#### Shooting:
- pctFTRate: The percentage of total field goal attempts that are Free Throws.
- pct3PRate: The percentage of total field goal attempts that are 3-pt shots.
- pctFG: Field Goal Percentage.
- pctFG3: Field Goal 3-pt Percentage.
- pctFG2: Field Goal 2-pt Percentage.
- pctEFG: Effective Field Goal Percentage. (FGM + (0.5 * 3PM)/FGA). This statistic adjusts for the fact that a 3-pt field goal is worth one more point than a regular 2-pt field goal.
- pctFT: Free Throw Percentage.
- fgmPerGame: Field Goal Made per Game.
- fgaPerGame: Field Goal Attempts per game
- fg3mPerGame: Field Goal Made (3-pt) per game.
- fg3aPerGame: Field Goal Attempts (3-pt) per game.
- fg2mPerGame: Field Goal Made (2-pt) per game.
- fg2aPerGame: Field Goal Attempts (2-pt) per game.
- ftmPerGame: Free Throws Made per game.
- ftaPerGame: Free Throw Attempts per game.
- ptsPerGame: Points per game.

#### Rebounds
- pctORB: Offensive Rebound Percentage (Rebound Rate).
- pctTRB: Total Rebound Percentage (Rebound Rate).
- pctDRB: Deffensive Rebound Percentage (Rebound Rate).
- orbPerGame: Offensive Rebounds per game.
- drbPerGame: Deffensive Rebounds per game.
- trbPerGame: Total Rebounds per game.

#### Passing
- pctAST: Assists Percentage.
- pctTOV: Turnover Percentage.
- astPerGame: Assists per game.
- tovPerGame: Turnovers per game.

#### Defense
- pctSTL: Steals percentage.
- pctBLK: Blocks percentage.
- stlPerGame: Steals per game.
- blkPerGame: Blocks per game.
- pfPerGame: Personal Fouls per game.

#### Efficiency stats
- ratioPER: Player Efficiency Rating
- ratioOWS: Offensive Win Shares.
- ratioDWS: Deffensive Win Shares.
- ratioWS: Win Shares.
- ratioWSPer48: Win Shares per 48 minutes.
- ratioOBPM: Offensive Box Plus/Minus.
- ratioDBPM: Deffensive Box Plus/Minus.
- ratioBPM: Box Plus/Minus.
- ratioVORP: Value Over Replacement Player.
- minutesPerGame: Minutes Per Game
- pctUSG: Usage Percentage. (100 * ((FGA + 0.44 * FTA + TOV) * (Tm MP / 5)) / (MP * (Tm FGA + 0.44 * Tm FTA + Tm TOV))

#### Other
- namePlayer: Name of player
- groupPosition: Player Position. Either Guard, Forward, or Center
- yearSeason: Season year
- countGames: Games played during season
- minutes: Minutes played during season
- isAllNBA: Did player make season's all NBA team?

```{r, message = FALSE, warning = FALSE, cache = TRUE}
# Gets player stats from 1984 to 2020.
players_post1984 <- bref_players_stats(seasons = 1984:2020, tables = c("advanced", "per_game"), include_all_nba = TRUE, return_message = FALSE)

# Filters by amount of time they played.
ggplot(players_post1984, mapping = aes(x = minutes)) + 
   geom_histogram() +
   geom_vline(xintercept = 600)
 players_post1984 <- filter(players_post1984, minutes > 600)

 # Variables to drop
dropping <-c("slugSeason", "idPlayerNBA", "namePlayerBREF", "urlPlayerBREF",
             "slugPlayerBREF", "urlPlayerThumbnail", "urlPlayerHeadshot",
             "slugPosition","urlPlayerPhoto", "urlPlayerStats",
             "urlPlayerActionPhoto","yearSeasonFirst",
             "countTeamsPlayerSeason","countGamesStarted",
             "isSeasonCurrent","slugPlayerSeason", "agePlayer",
             "slugTeamBREF", "isHOFPlayer", "slugTeamsBREF", 
             "countTeamsPlayerSeasonPerGame", "groupAllNBA",
             "numberAllNBATeam", "isAllNBA2", "isAllNBA1",
             "isAllNBA3" )
# Dropping variables
players_post1984 <- players_post1984[,!(names(players_post1984) %in% dropping)]

# Groups players by season, and normalizes by season.
for(year in 1984:2020){
  eval(parse(text = paste( "players_",year," <- filter(players_post1984, yearSeason == ",year,")", sep = "" )))
  for(i in 6:48){
    eval(parse(text = paste( "players_",year,"[,",i,"] <- normalize(players_",year,"[,",i,"],method = 'range', range = c(0,1))", sep = "")))
  }
}

# Ungroups players by season.
players_post1984_normalized <- players_1984
for(year in 1985:2020){
  eval(parse(text = paste( "players_post1984_normalized <- rbind(players_post1984_normalized, players_",year,")", sep = "")))
}
players_post1984_all_nba_normalized <- players_post1984_normalized %>% filter(isAllNBA == TRUE)
```

## Exploratory Data Analysis

```{r}
#some summary statistics
summarystatistics1984 <- players_post1984 %>%
  group_by(yearSeason, groupPosition) %>%
  summarize(meanpctFTRate = mean(pctFTRate),
            meanpct3PRate = mean(pct3PRate),
            meanpctFG = mean(pctFG),
            meanpctFG3 = mean(pctFG3),
            meanpctFG2 = mean(pctFG2),
            meanpctEFG = mean(pctEFG),
            meanpctFT = mean(pctFT),
            meanfgmPerGame = mean(fgmPerGame),
            meanfgaPerGame = mean(fgaPerGame),
            meanfg3mPerGame = mean(fg3mPerGame),
            meanfg3aPerGame = mean(fg3aPerGame),
            meanfg2mPerGame = mean(fg2mPerGame),
            meanfg2aPerGame = mean(fg2aPerGame),
            meanftmPerGame = mean(ftmPerGame),
            meanftaPerGame = mean(ftaPerGame),
            meanptsPerGame = mean(ptsPerGame),
            meanpctORB = mean(pctORB),
            meanpctTRB = mean(pctTRB),
            meanpctDRB = mean(pctDRB),
            meanorbPerGame = mean(orbPerGame),
            meandrbPerGame = mean (drbPerGame),
            meantrbPerGame = mean(trbPerGame),
            meanpctAST = mean(pctAST),
            meanpctTOV = mean(pctTOV),
            meanastPerGame = mean(astPerGame),
            meantovPerGame = mean(tovPerGame),
            meanpctSTL = mean(pctSTL),
            meanpctBLK = mean(pctBLK),
            meanstlPerGame = mean(stlPerGame),
            meanblkPerGame = mean(blkPerGame),
            meanpfPerGame = mean(pfPerGame),
            meanratioOWS = mean(ratioOWS),
            meanratioDWS = mean(ratioDWS),
            meanratioWS = mean(ratioWS),
            meanratioWSPer48 = mean(ratioWSPer48),
            meanratioOBPM = mean(ratioOBPM),
            meanratioDBPM = mean(ratioDBPM),
            meanratioBPM = mean(ratioBPM),
            meanratioVORP = mean(ratioVORP),
            sdpctFTRate = sd(pctFTRate),
            sdpct3PRate = sd(pct3PRate),
            sdpctFG = sd(pctFG),
            sdpctFG3 = sd(pctFG3),
            sdpctFG2 = sd(pctFG2),
            sdpctEFG = sd(pctEFG),
            sdpctFT = sd(pctFT),
            sdfgmPerGame = sd(fgmPerGame),
            sdfgaPerGame = sd(fgaPerGame),
            sdfg3mPerGame = sd(fg3mPerGame),
            sdfg3aPerGame = sd(fg3aPerGame),
            sdfg2mPerGame = sd(fg2mPerGame),
            sdfg2aPerGame = sd(fg2aPerGame),
            sdftmPerGame = sd(ftmPerGame),
            sdftaPerGame = sd(ftaPerGame),
            sdptsPerGame = sd(ptsPerGame),
            sdpctORB = sd(pctORB),
            sdpctTRB = sd(pctTRB),
            sdpctDRB = sd(pctDRB),
            sdorbPerGame = sd(orbPerGame),
            sddrbPerGame = sd (drbPerGame),
            sdtrbPerGame = sd(trbPerGame),
            sdpctSTL = sd(pctSTL),
            sdpctBLK = sd(pctBLK),
            sdstlPerGame = sd(stlPerGame),
            sdblkPerGame = sd(blkPerGame),
            sdpfPerGame = sd(pfPerGame),
            sdratioOWS = sd(ratioOWS),
            sdratioDWS = sd(ratioDWS),
            sdratioWS = sd(ratioWS),
            sdratioWSPer48 = sd(ratioWSPer48),
            sdratioOBPM = sd(ratioOBPM),
            sdratioDBPM = sd(ratioDBPM),
            sdratioBPM = sd(ratioBPM),
            sdratioVORP = sd(ratioVORP)
            )
head(summarystatistics1984)
ggplot(data = summarystatistics1984, mapping = aes(x = yearSeason, y = meanpctFG, color = groupPosition)) + 
  geom_line()
#taking a look at some stats over time, grouped by year and position
#fg pct
ggplot(data = summarystatistics1984, 
       mapping = aes(x = yearSeason, y = meanpctFG, color = groupPosition)) +  geom_line()
#3FGPct, see the introduction of 3 point line
ggplot(data = summarystatistics1984, 
       mapping = aes(x = yearSeason, y = meanpctFG3,
                     color = groupPosition)) +
  geom_line()
#meanFGM
ggplot(data = summarystatistics1984, 
       mapping = aes(x = yearSeason, y = meanfgmPerGame,
                     color = groupPosition)) +
  geom_line()
#meanFGA
ggplot(data = summarystatistics1984, 
       mapping = aes(x = yearSeason, y = meanfgaPerGame,
                     color = groupPosition)) +
  geom_line()
#mean3FGM 
ggplot(data = summarystatistics1984, 
       mapping = aes(x = yearSeason, y = meanfg3mPerGame,
                     color = groupPosition)) +
  geom_line()
#mean3FGA
ggplot(data = summarystatistics1984, 
       mapping = aes(x = yearSeason, y = meanfg3aPerGame, color = groupPosition)) +
  geom_line()
#meanFTA
ggplot(data = summarystatistics1984, 
       mapping = aes(x = yearSeason, y = meanftaPerGame, 
                     color = groupPosition)) +
  geom_line()
#meanFTM
ggplot(data = summarystatistics1984, 
       mapping = aes(x = yearSeason, y = meanftmPerGame,
                     color = groupPosition)) +
  geom_line()
#meanORB
ggplot(data = summarystatistics1984, 
       mapping = aes(x = yearSeason, y = meanorbPerGame,
                     color = groupPosition)) +
  geom_line()
#meanDRB
ggplot(data = summarystatistics1984, 
       mapping = aes(x = yearSeason, y = meandrbPerGame,
                     color = groupPosition)) +
  geom_line()
#meanAST
ggplot(data = summarystatistics1984, 
       mapping = aes(x = yearSeason, y = meanastPerGame, 
                     color = groupPosition)) +
  geom_line()
#meanSTL
ggplot(data = summarystatistics1984, 
       mapping = aes(x = yearSeason, y = meanstlPerGame,
                     color = groupPosition)) +
  geom_line()
#meanBLK
ggplot(data = summarystatistics1984, 
       mapping = aes(x = yearSeason, y = meanblkPerGame,
                     color = groupPosition)) +
  geom_line()
#meanTOV
ggplot(data = summarystatistics1984, 
       mapping = aes(x = yearSeason, y = meantovPerGame, 
                     color = groupPosition)) +
  geom_line()
#meanPTS
ggplot(data = summarystatistics1984, 
       mapping = aes(x = yearSeason, y = meanptsPerGame,
                     color = groupPosition)) +
  geom_line()

#also a correlation matrix plot
rcorr <- round(cor(players_post1984_normalized[,c(5:48)]),2)
corrplot(rcorr, method="color")
```


## Modeling
```{r, cache = TRUE}
pca <- prcomp(players_post1984_normalized[,-c(1:5, 49)])
d <- as.data.frame(pca$x)
pc1 <- pca$rotation[, 1]
pc1
pc2 <- pca$rotation[, 2]
pc2
pca3 <- pca$rotation[, 3]
d$namePlayer <- players_post1984_normalized$namePlayer
d$isAllNBA <- players_post1984_normalized$isAllNBA
d$yearSeason <- players_post1984_normalized$yearSeason

pcaplot <- ggplot(d, aes(x = PC1, y = PC2)) +
  geom_point(size = .5, alpha = .7) +
  ylab("Shooters vs Baseline") +
  xlab("Usage vs Non-Usage")
pcaplot

d1 <- tibble(PC = 1:43,
                PVE = pca$sdev^2 /
                  sum(pca$sdev^2))

ggplot(d1, aes(x = PC, y = PVE)) +
  geom_line() + 
  geom_point()

#some PCA exploration
pca_rotations <- data.frame(pca$rotation)
pca_rotations$variables <- rownames(pca_rotations)
PC1graph <- 
  ggplot(data = pca_rotations, mapping = aes(x = variables, y = PC1, 
                                          fill = variables)) +
  geom_col() +
  coord_flip() +
  theme(legend.position = "none") +
  labs(title = "Usage (-) vs Non-Usage (+)")
PC1graph

#note important characteristics: tov, pts, minutes, fgm, fga
PC2graph <- 
  ggplot(data = pca_rotations, mapping = aes(x = variables, y = PC2, 
                                          fill = variables)) +
  geom_col() +
  coord_flip() +
  theme(legend.position = "none")+
  labs(title = "Shooters vs Baseline")
PC2graph

#note important characteristics: (+) rebounding stats, (-) 3pt shooting, asts
PC3graph <- 
  ggplot(data = pca_rotations, mapping = aes(x = variables, y = PC3, 
                                          fill = variables)) +
  geom_col() +
  coord_flip() +
  theme(legend.position = "none")
PC3graph
#important characteristics: (+) turnovers, usage, field goals worth 2
#                            (-) True shooting, EFG, 3 pointers

#lets now create a plot with some players that we will recognize
currentplayersplot <- ggplot(d, aes(x = PC1, y = PC2)) +
  geom_point(size = .5, alpha = .1) +
  geom_text_repel(data = subset(d, isAllNBA == TRUE & 
                            yearSeason %in% 2017:2019 ),
            aes(label = namePlayer)) +
  geom_point(data = subset(d, isAllNBA == TRUE & 
                            yearSeason %in% 2017:2019), 
             mapping = aes(x = PC1, y = PC2),
             size = .5, alpha = .5, color = "purple2") +
  ylab("Shooters vs Baseline") +
  xlab("Usage vs Non-Usage")
currentplayersplot

famousplayersplot <-  ggplot(d, aes(x = PC1, y = PC2)) +
  geom_point(size = .5, alpha = .1) +
  geom_text_repel(data = subset(d, namePlayer %in% c("Lebron James", 
                                               "Michael Jordan",
                                               "Stephen Curry",
                                               "Kobe Bryant",
                                               "Tim Duncan",
                                               "Magic Johnson",
                                               "Kevin Durant")),
            aes(label = namePlayer)) +
  geom_point(color = "purple",
             data = subset(d, namePlayer %in% c("Lebron James", 
                                               "Michael Jordan",
                                               "Stephen Curry",
                                               "Kobe Bryant",
                                               "Tim Duncan",
                                               "Magic Johnson",
                                               "Kevin Durant"),
             alpha = .6)) +
  ylab("Shooters vs Baseline") +
  xlab("Usage vs Non-Usage")
famousplayersplot
#animated plot, allstars throughout the seasons
allnbaplayerspca <- d %>%
  filter(isAllNBA == TRUE)
#staticplot
staticplot <- ggplot(data = allnbaplayerspca, 
                     mapping = aes(x = PC1, y = PC2)) +
  geom_point(alpha = .8, color = "purple") + 
  geom_text_repel(data = allnbaplayerspca, 
                  mapping = aes(label = namePlayer)) +
  geom_point(data = d, mapping = aes(x = PC1, y = PC2), alpha = .1) +
  ylab("Shooters vs Baseline") +
  xlab("Usage vs Non-Usage")
staticplot


#byseason
byseasonplot <- staticplot + transition_time(yearSeason) +
  labs(title = "Season: {frame_time}")
animate(byseasonplot, renderer = gifski_renderer(), nframes = 37, fps = 2)

```

```{r, cache = TRUE}
n_pcas <- as_tibble(pca$x[,1:3])
#choosing k for kmeans
n_clusters_kmeans1<-kmeans(n_pcas, centers = 1, nstart = 20)
n_clusters_kmeans2<-kmeans(n_pcas, centers = 2, nstart = 20)
n_clusters_kmeans3<-kmeans(n_pcas, centers = 3, nstart = 20)
n_clusters_kmeans4<-kmeans(n_pcas, centers = 4, nstart = 20)
n_clusters_kmeans5<-kmeans(n_pcas, centers = 5, nstart = 20)
n_clusters_kmeans6<-kmeans(n_pcas, centers = 6, nstart = 20)
kmeansvariation <- data_frame("K" = 1:6,
                              "SS" = c(n_clusters_kmeans1$tot.withinss,
                                       n_clusters_kmeans2$tot.withinss,
                                       n_clusters_kmeans3$tot.withinss,
                                       n_clusters_kmeans4$tot.withinss,
                                       n_clusters_kmeans5$tot.withinss,
                                       n_clusters_kmeans6$tot.withinss))
kmeansvariationplot <- 
  ggplot(data = kmeansvariation, mapping = aes(x = K, y = SS)) +
  geom_line() +
  geom_point() +
  labs(title = "K means scree plot")
kmeansvariationplot
#does not seem to be a dinstinct elbow. We will decide on K=3, as it is
#near the middle of complete aggregation and no aggregation, and because
#generally, there seem to be three types of players


n_pcasdist <- dist(n_pcas)
distmatrix <- as.matrix(n_pcasdist)


n_clusters_hier<-hclust(dist(n_pcas))
n_clusters_hier_plot<-plot(n_clusters_hier,
                           xlab = "", ylab = "", sub = "", 
                           main = "Complete Linkage")
n_clusters_hier_plot

n_clusters_hier_cut <- cutree(n_clusters_hier, 3)

n_pcas_kmeans <- n_pcas %>% mutate(cluster = n_clusters_kmeans3$cluster)
n_pcas_hier <- n_pcas %>% mutate(cluster = n_clusters_hier_cut)

ggplot(n_pcas_kmeans, aes(x = PC1, y = PC2,
                          color = as.factor(cluster))) + 
  geom_point(alpha = .6) +
  labs(title = "kmeans clustering")
ggplot(n_pcas_hier, aes(x = PC1, y = PC2, 
                        color = as.factor(cluster))) +
  geom_point(alpha = .6) +
  labs(title = "hierarchical clustering")
##note, this appears different in the md doc than in my rstudio


#plot_ly(n_pcas_hier, x = ~PC1, y = ~PC2, z = ~PC3, color = ~cluster)

players_post1984_pca_clusters <- players_post1984 %>% 
  mutate(PC1 = pca$x[,1], PC2 = pca$x[,2], PC3 = pca$x[,3],
         hier_cluster = n_clusters_hier_cut, kmeans_cluster = n_clusters_kmeans3$cluster)

#what clusters do all-nba players end up in?
allnbaclusterssummary <- players_post1984_pca_clusters%>%
  filter(isAllNBA == TRUE) %>%
  group_by(hier_cluster) %>%
  summarize(n = n())


players_post1984_pca_clusters_year <- players_post1984_pca_clusters %>% 
  group_by(yearSeason, hier_cluster) %>% count

players_post1984_pca_year <- players_post1984_pca_clusters %>% 
  group_by(yearSeason) %>% count
players_post1984_pca_clusters_year_1 <- players_post1984_pca_clusters_year %>%
  filter(hier_cluster == 1)
players_post1984_pca_clusters_year_2 <- players_post1984_pca_clusters_year %>%
  filter(hier_cluster == 2)
players_post1984_pca_clusters_year_3 <- players_post1984_pca_clusters_year %>%
  filter(hier_cluster == 3)

cluster_proportions_1 = players_post1984_pca_clusters_year_1$n/players_post1984_pca_year$n
cluster_proportions_2 = players_post1984_pca_clusters_year_2$n/players_post1984_pca_year$n
cluster_proportions_3 = players_post1984_pca_clusters_year_3$n/players_post1984_pca_year$n
cluster_proportions <- tibble(year = c(1984:2020), one = cluster_proportions_1, two = cluster_proportions_2, three = cluster_proportions_3)

clusterdataframe <- data.frame(cluster1 = cluster_proportions_1,
                               cluster2 = cluster_proportions_2,
                               cluster3 = cluster_proportions_3,
                               year = c(1984:2020))
clusterdf <- data.frame(year = rep(c(1984:2020), 3),
                        proportion = c(cluster_proportions_1,
                                       cluster_proportions_2,
                                       cluster_proportions_3),
                        cluster = c(rep("Cluster 1", 37),
                                    rep("Cluster 2", 37),
                                    rep("Cluster 3", 37)))

#maybe use this graph instead?
ggplot(clusterdf, mapping = aes(x = year, y = proportion, color = cluster)) +
  geom_line()

#ggplot(cluster_proportions, aes(x = year, y = one), color = "red")+geom_line()+
 # geom_line(aes(x = year, y = two), color = "blue")+
  #geom_line(aes(x = year, y = three), color = "yellow")

players_post1984_pca_clusters_year
```

```{r}
pca_all_nba <- prcomp(players_post1984_all_nba_normalized[,-c(1:5, 49)])
d_all_nba <- as.data.frame(pca_all_nba$x)

ggplot(d_all_nba, aes(x = PC1, y = PC2)) +
  geom_point(size = .5, alpha = .7)

d2 <- tibble(PC = 1:43,
                PVE = pca$sdev^2 /
                  sum(pca$sdev^2))

d3 <- tibble(PC = 1:43,
             PVE = pca_all_nba$sdev^2 / sum(pca_all_nba$sdev^2))

ggplot(d2, aes(x = PC, y = PVE)) +
  geom_line() + 
  geom_point() +
  labs(title = "Skree Plot for NBA")


ggplot(d3, aes(x = PC, y = PVE)) +
  geom_line() + 
  geom_point() +
  labs(title = "Skree Plot for All-NBA")

pca_all_nba_rotations <- data.frame(pca_all_nba$rotation)
pca_all_nba_rotations$variables <- rownames(pca_all_nba_rotations)

PC1_all_nba_graph <- 
  ggplot(data = pca_all_nba_rotations, mapping = aes(x = variables, y = PC1, 
                                          fill = variables)) +
  geom_col()+
  coord_flip()+
  theme(legend.position = "none") +
  labs(title = "bigs(-) vs guards(+)")
PC1_all_nba_graph

PC2_all_nba_graph <- 
  ggplot(data = pca_all_nba_rotations, mapping = aes(x = variables, y = PC2, 
                                          fill = variables)) +
  geom_col()+
  coord_flip()+
  theme(legend.position = "none")
PC2_all_nba_graph

PC3_all_nba_graph <- 
  ggplot(data = pca_all_nba_rotations, mapping = aes(x = variables, y = PC3, 
                                          fill = variables)) +
  geom_col()+
  coord_flip()+
  theme(legend.position = "none")
PC3_all_nba_graph

PC4_all_nba_graph <- 
  ggplot(data = pca_all_nba_rotations, mapping = aes(x = variables, y = PC4, 
                                          fill = variables)) +
  geom_col()+
  coord_flip()+
  theme(legend.position = "none")
PC4_all_nba_graph
```





```{r}
n_pcas_all_nba <- as_tibble(pca_all_nba$x[,1:3])
n_clusters_kmeans_all_nba <-kmeans(n_pcas_all_nba, centers = 3)
n_clusters_hier_all_nba <-hclust(dist(n_pcas_all_nba))
n_clusters_hier_plot_all_nba<-plot(n_clusters_hier_all_nba)

n_clusters_hier_cut_all_nba <- cutree(n_clusters_hier_all_nba, 3)

n_pcas_kmeans_all_nba <- n_pcas_all_nba %>% mutate(cluster = n_clusters_kmeans_all_nba$cluster)
n_pcas_hier_all_nba <- n_pcas_all_nba %>% mutate(cluster = n_clusters_hier_cut_all_nba)

ggplot(n_pcas_kmeans_all_nba, aes(x = PC1, y = PC2,
                                  color = as.factor(cluster))) +
  geom_point() +
  labs(title = "all-nba kmeans clustering")
ggplot(n_pcas_hier_all_nba, aes(x = PC1, y = PC2, 
                                color = as.factor(cluster))) + 
  geom_point() +
  labs(title = "all-nba hierarchical clustering")

#plot_ly(n_pcas_hier_all_nba, x = ~PC1, y = ~PC2, z = ~PC3, color = ~cluster)

players_post1984_pca_clusters_all_nba <- players_post1984_all_nba_normalized %>% 
  mutate(PC1 = pca_all_nba$x[,1], PC2 = pca_all_nba$x[,2], PC3 = pca_all_nba$x[,3],
         hier_cluster = n_clusters_hier_cut_all_nba, kmeans_cluster = n_clusters_kmeans_all_nba$cluster)
players_post1984_pca_clusters_year_all_nba <- players_post1984_pca_clusters_all_nba %>% 
  group_by(yearSeason, hier_cluster) %>% count

players_post1984_pca_year_all_nba <- players_post1984_pca_clusters_all_nba %>% 
  group_by(yearSeason) %>% count
players_post1984_pca_clusters_year_1_all_nba <- players_post1984_pca_clusters_year_all_nba %>%
  filter(hier_cluster == 1)
players_post1984_pca_clusters_year_2_all_nba <- players_post1984_pca_clusters_year_all_nba %>%
  filter(hier_cluster == 2)
players_post1984_pca_clusters_year_3_all_nba <- players_post1984_pca_clusters_year_all_nba %>%
  filter(hier_cluster == 3)

cluster_proportions_1_all_nba = players_post1984_pca_clusters_year_1_all_nba$n/players_post1984_pca_year_all_nba$n
cluster_proportions_2_all_nba = players_post1984_pca_clusters_year_2_all_nba$n/players_post1984_pca_year_all_nba$n
cluster_proportions_3_all_nba = players_post1984_pca_clusters_year_3_all_nba$n/players_post1984_pca_year_all_nba$n

cluster_proportions_all_nba <- tibble(year = c(1984:2019), one = cluster_proportions_1_all_nba, two = cluster_proportions_2_all_nba, three = cluster_proportions_3_all_nba)

#Red: first cluster, Blue: second cluster, Green: third cluster
#ggplot(cluster_proportions_all_nba, aes(x = year, y = one))+geom_line(color = "red")+
 # geom_line(aes(x = year, y = two), color = "blue")+
  #geom_line(aes(x = year, y = three), color = "green") +
  #labs(y ="Proportion")
#new graph:
allnbaclusterdf <- data.frame(year = rep(c(1984:2019), 3),
                        proportion = c(cluster_proportions_1_all_nba,
                                       cluster_proportions_2_all_nba,
                                       cluster_proportions_3_all_nba),
                        cluster = c(rep("Cluster 1", 36),
                                    rep("Cluster 2", 36),
                                    rep("Cluster 3", 36)))
ggplot(allnbaclusterdf, mapping = aes(x = year, y = proportion, 
                                      color = cluster)) + 
  geom_line()

```


```{r}
#some more animated graphs
#counts
nbaclusters <- tibble(n_pcas_kmeans_all_nba$cluster, 
                          n_pcas_hier_all_nba$cluster,
                          allnbaplayerspca$yearSeason) %>%
  mutate(kmeanscluster=n_pcas_kmeans_all_nba$cluster ,
         hierarchicalcluster = n_pcas_hier_all_nba$cluster,
         yearSeason= allnbaplayerspca$yearSeason)
nbaclusters <- nbaclusters[,-c(1:3)]


clustercountkmeansallnba <- nbaclusters %>%
  group_by(yearSeason, kmeanscluster) %>%
  summarize(count = n())
clustercounthierchallnba <- nbaclusters %>%
  group_by(yearSeason, hierarchicalcluster) %>%
  summarize(count = n())


#allnba kmean and hierarchical clusters over time
kmeansclustergraphallnba <-
  ggplot(data = clustercountkmeansallnba, 
         mapping = aes(x = yearSeason, y = count, color = 
                         as.factor(kmeanscluster))) +
  geom_point() +
  ylim(-.5,7)
kmeansclustergraphallnba

hierarchicalgraphallnba <-
  ggplot(data = clustercounthierchallnba, 
         mapping = aes(x = yearSeason, y = count,
                       color = as.factor(hierarchicalcluster))) +
  geom_point() +
  ylim(-.5,7)
hierarchicalgraphallnba

#all-nba player's clusters over time
n_pcas_kmeans_all_nba <- n_pcas_kmeans_all_nba %>%
  mutate(yearSeason = allnbaplayerspca$yearSeason) 
n_pcas_hier_all_nba <- n_pcas_hier_all_nba %>%
    mutate(yearSeason = allnbaplayerspca$yearSeason) 
allnbakmeans <- 
  ggplot(n_pcas_kmeans_all_nba, aes(x = PC1, y = PC2, 
                                    color = as.factor(cluster)))+
  geom_point()+
  ylab("Shooters vs Baseline") +
  xlab("Usage vs Non-Usage") +
  labs(title = "All NBA teams over time - kmeans")
allnbakmeans
allnbahierch <- 
  ggplot(n_pcas_hier_all_nba, aes(x = PC1, y = PC2, 
                                  color = as.factor(cluster))) + 
  geom_point()+
  ylab("Shooters vs Baseline") +
  xlab("Usage vs Non-Usage") +
  labs(title = "All NBA teams over time - hierarchical")
allnbahierch  

byseasonplotkmeans  <- allnbakmeans + transition_time(yearSeason) +
  labs(title = "Season: {frame_time}") 
animate(byseasonplotkmeans, 
        renderer = gifski_renderer(), nframes = 36, fps = 2)
byseasonplothierch <- allnbahierch + transition_time(yearSeason) +
  labs(title = "Season: {frame_time}")
animate(byseasonplothierch, 
        renderer = gifski_renderer(), nframes = 36, fps = 2)

```



## Discussion

### Looking at PCA (Principal Component Analysis)

PC1 seems to be a case of an "Usage" (-) vs "Non-Usage" (+) battle.
In the negative side, we see characteristics such as ptsPerGame,
fgmPerGame, fg2mPerGame, fg2aPerGame, minutesPerGame. The other types
of variables do not tend to take on positive values, but the ones 
that are include:
pctBLK, pctSTL, pctTOV.  
Looking at PC2, I would describe this as "Shooters" (-) vs "Baseline" (+).
We see negative values for shooting characteristics, such as pct3PRate, 
fg3mPerGame, fg3aPerGame, and positive values for characteristics 
typical for tall, big players:pctDRB, pctTRB, blkPerGame.
PC3 is the distinction between "Outside the Arc" (-) and "Inside the Arc" (+)
Extreme negative values appear for pctTrueShooting, pctEFG, pct3PRate. On 
the positive side, we see tovPerGame, pctUSG, pctBLK, and fg2aPerGame. 

### Clustering
Using Hierarchical clustering, we were able to break down players into 3
"types": 
Cluster 1: very high usage
Cluster 2: medium usage
Cluster 3: low usage
These clusters were mainly focused around PC1; there was plenty of variance
in each cluster for PC2 and PC3.





###

## References

